Approximation as Differentiation

### 2025-07-10 main/language.tex
section numbering.
How Fam(MeetSLat×JoinSLatop) implements Sem.

\subsubsection{Syntax}
\label{sec:language:syntax}

@@

\subsubsection{Semantics}
\label{sec:language:semantics}

@@

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(+, 0)$ and exponentials
$\internalHom{X}{Y}$, with evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$,
plus the following additional structure:
\begin{enumerate}
\item a strong monad $(\Lift, \eta, \mu)$ with strength $\mathsf{st}_{X,Y}: X \times \Lift(Y) \to \Lift(X
\times Y)$
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}_{\PrimTy}$
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$
\item an endofunctor $\List: \Sem \to \Sem$, plus for any object $X$, morphisms $\nil: 1 \to \List(X)$ and
$\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to Y$ and
$f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$.
\end{enumerate}

@@

>>>>

\subsection{Syntax}
\label{sec:language:syntax}

@@

\subsection{Semantics}
\label{sec:language:semantics}

@@

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(+, 0)$ and exponentials
$\internalHom{X}{Y}$, with evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$,
plus the following additional structure:
\begin{enumerate}
\item a strong monad $(\Lift, \eta, \mu)$ with strength $\mathsf{st}_{X,Y}: X \times \Lift(Y) \to \Lift(X
\times Y)$
\item for each base type $\rho \in \PrimTy$, an object $\sem{\rho}_{\PrimTy}$
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$
\item an endofunctor $\List: \Sem \to \Sem$, plus for any object $X$, morphisms $\nil: 1 \to \List(X)$ and
$\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to Y$ and
$f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$.
\end{enumerate}

@@

\subsubsection{Interpretation for higher-order \GPS}

$\Fam(\MeetSLat \times \JoinSLat^\op)$ meets the requirements of $\Sem$: it is bicartesian closed
(\corref{mslat-jslat-bcc}), can model $\List(X)$ as an infinite coproduct
(\secref{models-of-total-gps:coproducts-and-products}), \roly{$\Lift$ monad, base types and operations.}

### 2025-07-10 main/models.tex
Standardise notation of List and L (lifting) functors.
How Fam(MeetSLat×JoinSLatop) implements Sem.

\subsubsection{Coproducts and Products}

@@

Using the infinitary coproducts and finite products, we can construct
a wide range of other useful semantic models of datatypes in
$\Fam(\cat{C})$. For example lists can be constructed as the coproduct
\begin{displaymath}
  \mathrm{List}\,X = \coprod_{n \in \mathbb{N}} X^n
\end{displaymath}
where $X^0 = 1$ (the terminal object) and $X^{n+1} = X \times X^n$.

@@

\begin{corollary}
  $\Fam(\MeetSLat \times \JoinSLat^\op)$ is cartesian closed and has
  all coproducts.
\end{corollary}

@@

On objects $X$ with trivial approximations, we have $L(X) \cong T(X)$,
but for more structured objects we get usage information at every level.

>>>>

\subsubsection{Coproducts and Products}
\label{sec:models-of-total-gps:coproducts-and-products}

@@

Using the infinitary coproducts and finite products, we can construct
a wide range of other useful semantic models of datatypes in
$\Fam(\cat{C})$. For example lists can be constructed as the coproduct
\begin{displaymath}
  \List(X) = \coprod_{n \in \mathbb{N}} X^n
\end{displaymath}
where $X^0 = 1$ (the terminal object) and $X^{n+1} = X \times X^n$.

@@

\begin{corollary}
  \label{cor:mslat-jslat-bcc}
  $\Fam(\MeetSLat \times \JoinSLat^\op)$ is cartesian closed and has
  all coproducts.
\end{corollary}

@@

On objects $X$ with trivial approximations, we have $\Lift(X) \cong T(X)$,
but for more structured objects we get usage information at every level.
